---
title: "Cogs 14B: Correlation and linear regression"
output:
  html_document:
    toc: yes
    toc_float: yes
---

*Sean Trott*

*February 5, 2020*
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(reshape)
```

# Introduction

The goal of this exercise is to give both a **conceptual** and **technical** introduction to correlations and linear regression.

# Correlation

A **correlation** is a measure between $[-1.0, 1.0]$ of the linear relationship between two quantitative variables $X$ and $Y$. In this tutorial, we'll be focusing on **Pearson's r** as a measure of correlation.

## Conceptual background: why compute correlations?

Put simply, a correlation allows you to infer: when $X$ goes up, does $Y$ tend to go up or down? A **positive** correlation ($r > 0$) means that when $X$ goes up, $Y$ also tends to go up; a **negative** correlation ($r < 0$) means that when $X$ goes up, $Y$ tends to go *down*. 

Furthermore, the **magnitude** of $r$ 